\documentclass[a4paper,10pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage[left=2cm,right=2cm,top=2cm,bottom=2cm]{geometry}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{setspace}

\begin{document}
\thispagestyle{empty}

\noindent
\begin{minipage}[t]{0.3\textwidth}
\textbf{\large CEUR}\\
\textbf{Workshop}\\
\textbf{Proceedings}\\
{\scriptsize \url{http://ceur-ws.org} \hfill ISSN 1613-0073}
\end{minipage}
\hfill
\begin{minipage}[t]{0.25\textwidth}
Version 1.07-ntp\\
2026-04-09
\end{minipage}
\hfill
\begin{minipage}[t]{0.4\textwidth}
CEUR Workshop Proceedings\\
\url{http://CEUR-WS.org}\\
GI e.V., Ahrstr. 45, 53175 Bonn, Germany
\end{minipage}

\vspace{0.6cm}

\begin{center}
{\Large \textbf{Author Agreement to Publish a Contribution as Open-Access on CEUR-WS.org}}
\end{center}

\vspace{0.3cm}

\noindent
Herewith I/we (the author(s) resp. the copyright holders) \textbf{agree} that my/our contribution:

\vspace{0.2cm}
\noindent
\textbf{MBART-50 Based Transfer Learning for Bidirectional MSL-Spanish Translation}
\vspace{0.2cm}

\noindent
authored by:

\vspace{0.2cm}
\noindent
\textbf{Miguel Bethencourt Mayedo}
\vspace{0.2cm}

\noindent
with corresponding author

\vspace{0.2cm}
\noindent
\begin{tabular}{@{}ll}
Name: & \textbf{Miguel Bethencourt Mayedo} \\
Affiliation: & \textbf{Universidad de Camag\"uey} \\
Address: & \textbf{Camag\"uey, Cuba} \\
Email: & \textbf{bethencourt.may@gmail.com}
\end{tabular}

\vspace{0.3cm}

\noindent
shall be made available as an open-access publication under the \textbf{Creative Commons License Attribution 4.0 International (CC BY 4.0)}, available at \url{https://creativecommons.org/licenses/by/4.0/legalcode}, and be published as part of the proceedings volume of the event

\vspace{0.2cm}
\noindent
Name and year of the event: \\
\textbf{IberLEF 2026}

\vspace{0.2cm}
\noindent
Editors of the proceedings (editors): \\
{\small \textbf{Alba Bonet-Jover, Jos\'e \'Angel G\'onzalez-Barba, Luis Chiruzzo, Alejandro Piad Morffis, Hugo Jair Escalante, Iker de la Iglesia, Albina Sarymsakova, Fazlourrahman Balouchzahi, Alba Mar\'ia M\'armol Romero, Horacio Saggion, Tania Gisela Alc\'antara Medina, Ansel Yoan Rodr\'iguez Gonz\'alez, Mar\'ia Victoria Cantero Romero, Jos\'e Antonio Garc\'ia-D\'iaz, \'Alvaro Rodrigo, Miguel \'Angel \'Alvarez Carmona, Vicent Ahuir, Luis Israel Ramos P\'erez, Niels Mart\'inez Guevara}}

\vspace{0.3cm}

\noindent
{\footnotesize I/we agree that my/our contribution is made available publicly under the aforementioned license on the servers of CEUR Workshop Proceedings (CEUR-WS). I/we grant the editors, GI e.V., CEUR-WS, and its archiving partners the non-exclusive and irrevocable \textbf{right to archive} my/our contribution and \textbf{to make it accessible} (online and free of charge) for \textbf{public distribution}. This granted right extends to any associated metadata of my/our contribution. Specifically, I/we license the associated metadata under a Creative Commons CC0 1.0 Universal license (public domain). I/we agree that our author names and affiliations is part of the associated metadata and may be stored on the servers of CEUR-WS and made available under the CC0 license. I/we acknowledge that the editors hold the copyright for the proceedings volume of the aforementioned event as the official collection of contributions to the event.\par}

\vspace{0.2cm}
\noindent
{\footnotesize \textbf{I/we have not included any copyrighted third-party material such as figures, code, data sets and others in the contribution to be published. I/we have included a declaration on the use of Generative AI (if any) in the paper text.}\par}

\vspace{0.2cm}
\noindent
{\footnotesize I/we warrant that my/our contribution (including any accompanying material such as data sets) does not infringe any rights of third parties, for example trademark rights, privacy rights, and intellectual property rights. I/we understand that I/we retain the copyright to my/our contribution. I/we understand that the dedication of my/our contribution under the CC BY 4.0 license is irrevocable.\par}

\noindent
{\footnotesize I/we understand and agree that the \textbf{full responsibility/liability} for the content of the contribution rests upon me/us as the authors of the contribution. I/we \textbf{release and relieve} the aforementioned editors, GI e.V., individuals providing the CEUR-WS service, and the archiving partners of CEUR-WS of \textbf{liability caused by ordinary/simple negligence} in the publication or archiving of my/our aforementioned contribution via the servers used by CEUR-WS.\par}

\noindent
{\footnotesize I/we have read the conditions of the Creative Commons License Attribution 4.0 International (CC BY 4.0), and agree to apply this license to my/our contribution.\par}

\vspace{1.0cm}

\noindent Camag\"uey, Cuba, 9 de junio de 2026 \hspace{3.2cm} \rule{0.4\textwidth}{0.4pt} \\
Location, Date, Signature of the corresponding author representing all authors \\
\textbf{(Signature must be handwritten with a pen on paper)}

\end{document}
